---
title: "Statistical Inference Project, Part 1"
author: "Hoang Tam Vo"
output: pdf_document
---

<!-- The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also also 1/lambda. Set lambda = 0.2 for all of the simulations. In this simulation, you will investigate the distribution of averages of 40 exponential(0.2)s. Note that you will need to do a thousand or so simulated averages of 40 exponentials.
Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponential(0.2)s. -->

In this part of the project, I study the exponential distribution in R and compare it with the Central Limit Theorem. Specifically, in R, the exponential distribution can be simulated with `rexp(n, lambda)` where `lambda` is the rate parameter. 
<!--The mean of exponential distribution is `1/lambda` and the standard deviation is also `1/lambda`.--> 
I illustrates via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponential(0.2)s.

# Simulation
In this study, `lambda` is set to 0.2 for all of simulations. A thousand of simulations are done and  the distribution of averages of 40 exponentials are investigated.

```{r}
lambda = 0.2
n = 40
nsms = 1:1000
set.seed(26042015)
means <- data.frame(x = sapply(nsms, function(x) {mean(rexp(n, lambda))}))
head(means)
```

# Questions 1: Sample Mean versus Theoretical Mean

The average sample mean of 1000 simulations of 40 randomly sampled exponential distributions:
```{r}
mean(means$x)
```

The expected mean of an exponential distribution of rate `lambda` is:
```{r}
expectedMean <- 1/lambda
expectedMean
```

As can be seen, the expected mean and the avarage sample mean are very close.

# Questions 2: Sample Variance versus Theoretical Variance

The standard deviation of the average sample mean of 1000 simulations of 40 randomly sampled exponential distribution:

```{r}
sd(means$x)
```


The variance of the average sample mean of 1000 simulations of 40 randomly sampled exponential distribution:

```{r}
var(means$x)
```

The expected standard deviation of an exponential distribution of rate `lambda` is:
```{r}
(1/lambda)/sqrt(n)
```


<!-- The variance of an exponential distribution of rate `lambda` is:-->
The expected variance of an exponential distribution of rate `lambda` is:
```{r}
((1/lambda)/sqrt(n))^2
```

As can be seen, the standard deviations are very close, while variances are also pretty close as variance is the square of the standard deviation.

# Question 3: Show that the distribution is approximately normal

A histogram plot of the means of the 1000 simulations of rexp(n, lambda) is shown in the below figure. In addtion, it is overlaid with a normal distribution with mean 5 and standard deviation 0.781. 
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
ggplot(data = means, aes(x = x)) + 
    geom_histogram(aes(y=..density..), fill = I('#00e6fa'), 
                   binwidth = 0.20, color = I('black')) +
    stat_function(fun = dnorm, arg = list(mean = 5, sd = sd(means$x)))
     
```

As can be seen, the distribution of our simulations is approximately normal.



