---
title: "Statistical Inference Project, Part 2"
author: "Hoang Tam Vo"
output: pdf_document
---

# Overview


In this part of the project, I analyze the ToothGrowth data in the R datasets package. Specifically, I load the ToothGrowth data and perform some basic exploratory data analyses as well as provide a basic summary of the data. Further, confidence intervals and hypothesis tests are used to compare tooth growth by supp and dose. Finally, conclusions and the assumptions needed are discussed. 

# Questions 1: Load the ToothGrowth data and perform some basic exploratory data analyses


```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(datasets)
data(ToothGrowth)
str(ToothGrowth)
ToothGrowth[c(1:8),]
```

```{r}
plot <- ggplot(ToothGrowth, aes(x=factor(dose), y=len, fill=factor(dose)))

plot + geom_boxplot(notch=F) +
     facet_grid(.~supp) +
     scale_x_discrete("Dosage (milligram)") +   
     scale_y_continuous("Length of teeth")  + 
     ggtitle("Exploratory data analyses")
```


# Questions 2: Provide a basic summary of the data


```{r}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
summary(ToothGrowth)
```

```{r}
table(ToothGrowth$dose, ToothGrowth$supp)

```


# Question 3: Use confidence intervals and hypothesis tests to compare tooth growth by supp and dose

```{r}
test1 <- t.test(len~supp, paired=F, var.equal=T, data=ToothGrowth)
test2 <- t.test(len~supp, paired=F, var.equal=F, data=ToothGrowth)
result <- data.frame("p-value"=c(test1$p.value, test2$p.value),
                          "Low Confidence"=c(test1$conf[1],test2$conf[1]),
                          "High Confidence"=c(test1$conf[2],test2$conf[2]),
                          row.names=c("Equal","Unequal"))
result
```


# Question 4: State  conclusions and the assumptions needed

Given the above analysis, several conclusions can be drawn: 

1. Orange juice (OJ) and vitamin C (VC) have different effect on the growth of the tooth.

2. There is a difference in the tooth growth when the doses are larger.

These analysis and conclusions are based on the followwing assumptions:

* The sample data are repesentative for the entire population.

* The distribution of the data is normal.

